CAP in practice: HBase and MongoDB

### 2014-08-22 Thesis - 2/Paper/paper.tex
Ingeleverde versie

\usepackage{array}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\author{Thomas Uyttendaele}

@@

\begin{abstract}
This papers introduced a new, quantified method to test distributed database system in regard to their behaviour to availability in case of a failing node and consistency of data from the users perspective. This method is tested on HBase and MongoDB, two consistent and partition tolerant systems, with the most important outcomes included. 
\end{abstract}

@@

New online services and more online users means more data, data and load a single server can't handle. More and more database systems are distributed, for higher availability in case of an unexpected crash but also for horizontal distribution: the data of a single database is spread over different systems. These new services have different requirements; e.g.. an update doesn't need to be visible for all users immediately, it can take time, this behaviour is called eventual consistency.

Over the past years, many new systems have been build on this wave of changes, categorized as NoSQL. Some applications contain a high volume of data, have the need for consistent data and being able to work in a highly distributed environment, this are the CP systems in the CAP Theorem. Two examples that offer these guarantees are HBase and MongoDB. They greatly differ in supported queries on their system, but what happens if you only use the bare essentials and compare their behaviour in a distributed environment going from expected shut downs of instances towards crashes and network partitions?

In this article, a comparison of both this systems on these 3 behaviours will be made. In chapter \ref{sec:CAPTheorem} a brief overview of the CAP Theorem is given, chapter \ref{sec:hbaseandmongodb} discusses HBase and MongoDB on paper. Chapter \ref{sec:testmethod} gives an overview of the used test method and chapter \ref{sec:result} presents the results. The future work is presented in chapter \ref{sec:futurework}, an overview of related work is given in chapter \ref{sec:related work} and a conclusion is made in chapter \ref{sec:conclusion}.

\section{The CAP Theorem\cite{brewer2000towards}\cite{brewer2012cap}}\label{sec:CAPTheorem}
The CAP Theorem was introduced by E. Brewer \cite{brewer2000towards}  in 2000 and discusses 3 properties of which each network shared-data system can guarantee at most 2: (definitions based on \cite{brewer2012cap})
\begin{itemize}
\item (Strict)\textbf{C}onsistency: The system acts like there is only single storage 
\item High \textbf{a}vailability: The system is available (for updates)
\item \textbf{P}artition tolerance: A split in the network let the different partitions still act as a single system. 
\end{itemize}
Designers used this model to explain their design decisions, others used it to compare different system and sometimes it was misused. As E. Brewer explains 12 year after the launch, the "2 of 3" can be misleading.

One of the reasons is that there exist several types of consistency, partition tolerance and different availability guarantees. The choice for the properties needs to be made several times in the different subsystems and the end solution is not black or white.

At first glance, it also looks like partition tolerance has to be implemented and therefore consistency and availability needs to be given up. In the practice, partition splits are only rare and therefore both consistency and availability can be allowed most of the time.

@@

These systems can be build around the 2 phase commit and have cache invalidation protocols. Examples of this types are the typical relational databases roll out in clusters.

\subsection{CP: Consistency and partition tolerance}
A consistent system with partition tolerance will provide all the time the last data, even in the case of network splits. This comes with the loss of the availability of all nodes all the time.

@@

\subsection{AP: Availability and partition tolerance}
In a highly available systems with partition tolerance, is it possible to read inconsistent data. As read and write operations are still allowed when there are different partitions, it is possible that the database has other content depending on the used node. When the split is dissolved, a need for manual conflict resolution can be needed. In case a record is adapted in both partitions, the user will need to choose the correct version.

@@

\subsection{HBase}
HBase\cite{hbase-doc} is an open-sourced, distributed, versioned database designed after Google's BigTable \cite{chang2008bigtable}. HBase relies on Zookeeper for the distributed task coordination and the persistent storage can be done on the local hard disk, Hadoop Distributed File System or Amazon S3. In this article is chosen for Hadoop.

HBase nodes exists out of HMasters and HRegionServers, the coordination of the system is done by one HMaster, the handling of data is done by the HRegionServer. To store the data, multiple Hadoop datanode instance should be deployed for data storage, preferable one on each HRegionServer. The data will be replicated to a configurable amount of other nodes, default modus is 3. The data is stored in a table, which is split in one or more regions. A region is leased to a given HRegionServer for a defined time. During this time, only this server will provide the data of the region to the different users. This way the consistency of data can be guaranteed because there is for each record only a single system responsible. Consistency on a single record is provided by a readers/writer lock on a single record for the according queries, this way there is a guarantee to atomicity on a single record, the full procedure is explained by Lars Hofhansl\cite{hbase-acid}.

@@

In HBase, a node will be able to answer to requests if the node is present in the data serving partition. Only in rare cases that all data copies of Hadoop are stored in unavailable servers, the data will be unreachable. The nodes not in the data serving partition, will be unable to complete any requests.

@@

MongoDB provides data replication and data distribution, the first is done by grouping different MongoDB servers into a ReplicaSet, the second is done by grouping different of these ReplicaSets.

A ReplicaSet exists out of different MongoDB servers whom work as a master/slave configuration. A master is a primary and a slave is called a secondary. The primary is responsible for the write actions, by default a query will succeed once it has a confirmation that the write has been executed on the primary. The read operation will go by default on the primary as well. Both query methods are configurable to give other guarantees, for a write operation there are multiple \textit{write concerns}, it is possible to wait till it has written on hard disc or a number of secondary servers, however all need a primary. For read operation there are multiple \textit{read preferences}, it is possible to read from a secondary or the closest server. \\ 
In the default configuration, MongoDB provides a consistency guarantee.

@@

The state of the different members of a ReplicaSet is maintained by a heartbeat system: a server is marked as offline if no beat has been received for 10 seconds. In case the primary goes offline, election will be started to re-elect a new one. In other words, a primary has a lease of 10 seconds. This value of 10 seconds is non-configurable.

@@

First of all, they differ in their handling of partition tolerance, in HBase there are dedicated management servers (HMaster and Zookeeper) to distribute the responsibilities of regions and if the management servers are in a partition with a minority of the data servers, data will be not available. In MongoDB the management of a ReplicaSet is done internally with as result that the write queries will be available in the partition with the majority of the data servers.

@@

\section{Test method}\label{sec:testmethod}
To test the behaviour of database systems towards consistency and availability, the Yahoo Cloud Serving Benchmarking (YCSB \cite{cooper2010benchmarking}) has been extended with event support for availability and reader-writers for consistency.

@@

\subsection{Event support}
The implementation of event support is integrated so it supports the execution of UNIX commandos at specified moments, the execution time and result code is logged.

Before the test are 300 000 records stored in the database to enable the sharding in all database softwares. In these tests were 3 kinds of tests executed of which each one took 900 second. The first action takes place at 300 seconds, the second at 600 seconds.
\begin{itemize}
\item Graceful shut down of a data service and restart of the service
\item Hard kill of the data service and a restart of the service
\item Blocking of all network traffic from and to a server and the allowance of all network traffic
\end{itemize}

Each element tests the behaviour of the system under different circumstances, the first two checks what happens in case of a respectively planned and unexpected shut down, the latter check the behaviour in case the network fails.

The tests are executed on each of the data nodes of HBase and MongoDB, caching and buffering on the client side are disabled in both tests.

\subsection{Consistency support}
To implement consistency support, the YCSB software is extended with an extra workload. A graphical representation of an example workload is shown in figure \ref{fig:consistence-method}. 
\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{../img/Consistency-test-period}
\caption{Example of consistency support for one write period with 1 writer and 4 readers}
\label{fig:consistence-method}
\end{figure}
This workload exists out of 1 writer and a user-defined amount of readers.  Each writer will inserts or updates a value in the database with at a user-defined pace. The readers will read the record till they read the last written data element, each reader reads in a period defined by the user. The different readers are scheduled uniformly within the reading period.

All this data is logged and gives possibilities to analyse when a record is visible for different users, compared to the time the queries where started or ended.

Before the test, there are 30 000 records stored in the database, each tests takes 500 seconds and results start to be gathered after 30 seconds. In the tests is chosen to write every 0.5s and read with 5 readers every 10ms for MongoDB, for HBase there are 10 readers who read every 30ms.

@@

YCSB was deployed on a single instance and used to calibrate the load on both systems. An individual record has 10 fields which each field a size of 100 bytes. The workload existed out of 20\% inserts and updates, 40\% selects and 20\% scans of an uniform spread between 1 and 100. The requests are spread according to Zipfian\footnote{Some record are popular, others rare to be used.}  The basic load for HBase is an average of 600 queries/second spread over 50 threads, for MongoDB there are 15 threads with a total of 200 queries/second.

@@

In case of a stop of a HBase server in this configuration, the queries will halt till the lease for the region has been expired, this can take between 0 seconds and 180, depending on the moment of the action. In case of a graceful stop of the service, the service will release the lease and a faster handover is possible. During the tests it happened that with a hard stop of the service, the queries are impossible on the connections till the server is brought back online. In this scenario it was needed to manually dis- en reconnect to the database. A reason why this happens here is not known.

For MongoDB, there is no difference between the graceful or hard stop of an instance; in case it was a secondary, no influences on the latency will be seen. In case a primary was stopped, the data will be in both cases temporarily unavailable for a few seconds. It seems that in a hard stop, there is still a messaging of the shut down to the other nodes. In case of a network interruption, the results are varied, going from no influence, a temporary unavailability for a few seconds, or no completed queries the whole time the server is down. When in the last case, all connections are reinitialised, the queries can happen, when this is not done the table seems empty.

@@

\subsection{Consistency}
In consistency HBase and MongoDB have a different approach, in MongoDB it is possible to configure the read and write queries, in HBase there is the choice to use cache at client side. In the tests the cache is disabled but the different options of MongoDB are tested.

HBase: If a record is read while a write query is creating or updating the value, the read query will wait till the completion of the write query and return immediately the correct value. If a read query is send too soon, it will return the old data, but each query returned after the completion of a write query, will return the new data. A graphical representation of this can be seen in figure \ref{fig:consistentie-hbase-voorbeeld}. The stop time of the reader follows the one of the writer, together with extra information, it can be concluded that this is coming from the blocking till the write query has finished.

@@

MongoDB: A read query can return the old or new data before the write query has been completed. However a readers/writer lock mechanism is used, it seems that after releasing the lock, the write query still has to execute extra steps. When reading from a secondary or the nearest server, there will be monotone read consistency, except if the driver of MongoDB selects another server between the reads, as there is no guarantee that the other servers already have the value. The process of finding the nearest server happens periodically in the background. \\
From the test results, the writer configuration has no influence on the speed the replication will have, as the read queries get the same consistency window based on the start moment. The difference is the guarantee a user has when a write query has finished.

An overview of the consistency windows is shown in table \ref{table:consistentie-hbase-correct} and \ref{table:consistentie-mongodb-correct} for respectively HBase and MongoDB.

\begin{table}
\centering
\begin{tabular}{l| l l}
	\textbf{Reader} & \textbf{Start time (ms)} & \textbf{Percentage correct read} \\
	\hline
	1 &  0 ms & 2.6\%\\
	2 &  3 ms & 68\%\\
	3 &  6 ms & 90\%\\
	4 &  9 ms & 93\%\\
	5 &  12 ms & 94\%\\
	6 &  15 ms & 96\%\\
	7 &  18 ms & 96\%\\
	8 &  21 ms & 96\%\\
	9 &  24 ms & 97\%\\
	10 & 27 ms & 97\% 
\end{tabular}
\caption{Consistency: Percentage of the queries started at the given time moment that will read the newest inserted data. The average latency in a random read transaction is around 6ms.}
\label{table:consistentie-hbase-correct}
\end{table}

\begin{table*}
\centering
\begin{tabular}{l | l l l l}
& nearest & primary & primary- \newline preferred & secondary \\ 
  \hline
safe & 28, 69, 89, 91, 92 & 80, 98, 98, 99, 99 & 74, 99, 99, 99, 99 & 0, 65, 83, 85, 88 \\ 
  normal & 24, 68, 87, 89, 92 & 72, 99, 100, 100, 100 & 75, 98, 98, 98, 98 & 0, 69, 85, 89, 92 \\ 
  fsync\_safe & 28, 73, 87, 90, 90 & 68, 96, 98, 98, 98 & 78, 97, 98, 98, 98 & 0, 66, 80, 85, 86 \\ 
  replicas\_safe & 24, 74, 87, 88, 91 & 75, 98, 99, 99, 99 & 79, 98, 98, 98, 98 & 1, 67, 84, 87, 89 \\ 
  majority & 26, 77, 91, 91, 92 & 73, 98, 99, 99, 99 & 77, 99, 99, 99, 100 & 0, 61, 82, 85, 89 \\ 
\end{tabular}
	\caption{Consistency: Percentage of the queries started at 0, 2, 4 and 8ms  that will read the newest inserted data with the each row for a write configuration and a column for the read configuration. The average latency on a read transaction is for all read queries around 1ms. }
	\label{table:consistentie-mongodb-correct}
\end{table*}

\section{Future work}\label{sec:futurework}
The test run on the different systems are a start to have a quantified approach on the consistency and availability of the different systems. However multiple extensions can be made to this test method, next to testing more systems.

@@

Secondly, the third element of CAP can be included: Partition tolerance. Right now the connection nodes are chosen when setting up the test, an extension could be to test all connections and see what happens in the case of a partition tolerance. How is the availability of each node, how long is a primary in MongoDB still accepting queries however it is cut off?

@@

At last, the test parameters could be adapted, what happens with a different network infrastructure when the network distance is not equal any more, what happens with a higher and lower load? It could be possible in the end to have a mathematical formula which could predict the average consistency window for example.

\section{Related work}\label{sec:related work}
The research towards database systems and their consistency guaranties is rare to have a measured approach. In recent paper (February 2014), Golab et al. states that there is only a limited amount of research done towards eventual consistency \cite{golab2014eventually}. They present a new view on consistency, were already some research is done towards active analyse (how long before the data is replicated to all nodes), the amount of passive analyse is limited (what do the users see). Compared to the consistency results from this paper, both are discussed: as well the delays before the data is present everywhere but also on the acting of the specific systems, in example the behaviour of HBase.

Another extension of YCSB called YCSB++\cite{patil2011ycsb++}, provides more logging information on all systems in the first place, but they also test the consistency of HBase in regards of the client caches. The reasoning for this is that it is the standard run configuration of HBase, but the submitting of the client cache towards the system depends not only on the time, but also on the amount of traffic that is being submitted. In the study they compare different cache sizes and this shows already a difference in time. Furthermore, it is possible to disable this caching in case there are records in the need of this strict consistency.

@@

In the availability of the systems, HBase is in the standard configuration longer unavailable but with a change in the session time out, they can be made more the same. However both have a strange behaviour of not returning any results, in the case of HBase this is for a hard stop, in the case of MongoDB it is for a network partition. This doesn't happen in the other situations and in case of reconnecting to the database, the problems are solved. 
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

>>>>

\usepackage{array}

\usepackage{tikz}
\usepackage{pgfplots}

\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\author{Thomas Uyttendaele (thomas.uyttendaele@student.kuleuven.be) \\ 
Departement Computerwetenschappen, KULeuven, Belgi\"e}

@@

\begin{abstract}
This papers introduces a new, quantified method to test distributed database system in regard to their behaviour to availability in case of a failing node and consistency of data from the users perspective. This method is tested on HBase and MongoDB, two consistent and partition tolerant systems, some interesting outcomes are described in this paper. 
\end{abstract}

@@

New online services and more online users means more data, data and load a single server can't handle. Database systems are transformed towards a more distributed approach, for higher availability in case of an unexpected crash but also for horizontal scalability: the data of a single database is spread over different systems. Several new software have other requirements on the database than a few years ago; e.g. a change in a record doesn't need to be read correctly by other users immediately, this can happen in time. This behaviour is called eventual consistency.

Over the past years, many new systems have been build on this wave of changes, categorized as NoSQL. Some applications contain a high volume of data, have the need for consistent data and need to work in a highly distributed environment, this are the CP systems in the CAP Theorem. Two examples that offer these guarantees are HBase and MongoDB. They greatly differ in supported queries on their system. But what happens if you only use the bare essentials and compare their behaviour in a distributed environment going from expected shut downs of instances towards crashes and network partitions?

In this article, a comparison of this systems will be described. In chapter \ref{sec:CAPTheorem} a brief overview of the CAP Theorem is given, chapter \ref{sec:hbaseandmongodb} discusses HBase and MongoDB based on the literature. Chapter \ref{sec:testmethod} gives an overview of the used test method and chapter \ref{sec:result} presents the results. The future work is presented in chapter \ref{sec:futurework}, an overview of related work is given in chapter \ref{sec:related work} and a conclusion is formulated in chapter \ref{sec:conclusion}.

\section{The CAP Theorem\cite{brewer2000towards}\cite{brewer2012cap}}\label{sec:CAPTheorem}
The CAP Theorem was introduced by E. Brewer \cite{brewer2000towards}  in 2000 and discusses three properties of which each network shared-data system can guarantee at most two: (definitions based on \cite{brewer2012cap})
\begin{itemize}
\item (Strict)\textbf{C}onsistency: The system acts like there is only single storage 
\item High \textbf{a}vailability: The system is available (for updates)
\item \textbf{P}artition tolerance: A split in the network let the different partitions still act as a single system. 
\end{itemize}
Designers used this model to explain their design decisions, others used it to compare different system and sometimes it was misused. As E. Brewer explains 12 year after the launch, the "2 of 3" can be misleading.

One of the reasons is that there exist several types of consistency, partition tolerance and there are different availability guarantees. The choice for the properties needs to be made several times in the different subsystems and the end solution is not black or white.

At first glance, it also looks like partition tolerance has to be implemented and therefore consistency or availability needs to be given up. In the practice, partition splits are only rare and therefore both consistency and availability can be allowed most of the time.

@@

These systems can be build based on the two phase commit and have cache invalidation protocols. Examples of this types are the typical relational databases roll out in clusters.

\subsection{CP: Consistency and partition tolerance}
A consistent system with partition tolerance will provide all the time the last data, even in the case of network splits. This comes with the loss of the availability of some nodes during the partition time.

@@

\subsection{AP: Availability and partition tolerance}
In a highly available systems with partition tolerance. It is possible to read inconsistent data. As read and write operations are still allowed when there are different partitions, it is possible that the database has other content depending on the used node. When the split is dissolved, a need for manual conflict resolution can be needed. In case a record is adapted in both partitions, the user will need to choose the correct version.

@@

\subsection{HBase}
HBase\cite{hbase-doc} is an open-sourced, distributed, versioned database designed after Google's BigTable \cite{chang2008bigtable}. HBase relies on Zookeeper for the distributed task coordination. The persistent storage can be done on the local hard disk, Hadoop Distributed File System or Amazon S3. In this article is chosen for Hadoop.

HBase nodes exists out of HMasters and HRegionServers, the coordination of the system is done by one HMaster, the handling of data is done by the HRegionServer. Multiple Hadoop datanode instance should be deployed for data storage, preferable one on each HRegionServer. The data will be replicated to a configurable amount of other nodes, default modus is 3. The data is stored in a table, which is split in one or more regions. A region is leased to a given HRegionServer for a defined time. During this time, only this server will provide the data of the region to the different users. This way the consistency of data is guaranteed due to the fact that there is a single system responsible for the storage of a record.  Consistency on a single record is provided by a readers/writer lock on a single record for the according queries, this way there is a guarantee to atomicity on a single record, the full procedure is explained by Lars Hofhansl\cite{hbase-acid}.

@@

In HBase, a node will be able to reply to requests if the node is present in the data serving partition. Only in rare cases that all data copies of Hadoop are stored in unavailable servers, the data will be unreachable. The nodes that are not in the data serving partition, will be unable to complete any requests.

@@

MongoDB provides data replication and data distribution. The first is done by grouping different MongoDB servers into a ReplicaSet. The second is done by grouping different of these ReplicaSets and dividing the data between them.

A ReplicaSet exists out of different MongoDB servers whom work as master and slaves. A master is a primary and a slave is called a secondary. The primary is responsible for the write actions, by default a query will succeed once it has a confirmation that the write has been executed on the primary. The read operation will go by default on the primary as well. Both query methods are configurable to give other guarantees. For a write operation there are multiple \textit{write concerns}, it is possible to wait till it has written on hard disc or a number of secondary servers, however all need a primary. For read operation there are multiple \textit{read preferences}, it is possible to read from a secondary or the closest server. \\ 
In the default configuration, MongoDB provides a consistency guarantee.

@@

The liveness of the different members of a ReplicaSet is maintained by a heartbeat system: a server is marked as offline if no beat has been received for 10 seconds. In case the primary goes offline, election will be started to re-elect a new one. In other words, a primary has a lease of 10 seconds. This value of 10 seconds is non-configurable.

@@

First of all, they differ in their handling of partition tolerance. HBase has dedicated management servers (HMaster and Zookeeper) to distribute the responsibilities of regions and if the management servers are in a partition with a minority of the data servers, data will be not available. In MongoDB the management of a ReplicaSet is done internally with as result that the write queries will be available in the partition with the majority of the data servers.

@@

\section{Test method}\label{sec:testmethod}
To test the behaviour of database systems towards consistency and availability, the Yahoo Cloud Serving Benchmarking (YCSB \cite{cooper2010benchmarking}) has been extended with event support for availability and reader-writer support for consistency.

@@

\subsection{Event support}
The implementation of event support is integrated in YCSB so it supports the execution of UNIX commandos at specified moments. The execution time and result code is logged.

Before the test are 300 000 records stored in the database to enable the sharding in all database softwares. These tests contains 3 kinds of tests, of which an individual test takes 900 second. The first action takes place at 300 seconds, the second at 600 seconds.
\begin{itemize}
\item (1) Graceful shut down of a data service and (2) restart of the service
\item (1) Hard kill of the data service and (2) a restart of the service
\item (1) Blocking of all network traffic from and to a server and (2) the allowance of all network traffic
\end{itemize}

Each element tests the behaviour of the system under different circumstances, the first two checks what happens in case of a respectively planned and unexpected shut down, the latter check the behaviour in case the network fails or a server crashes.

The tests are executed on each of the data nodes of HBase and MongoDB.Caching and buffering on the client side are disabled in both tests.

\subsection{Consistency support}
To implement consistency support, the YCSB software is extended with an extra workload. A graphical representation of an example workload is shown in figure \ref{fig:consistence-method}. 
\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{../img/Consistency-test-period}
\caption{Example of consistency support for one write period with 1 writer and 4 readers}
\label{fig:consistence-method}
\end{figure}
This workload exists out of 1 writer and a user-defined amount of readers.  Each writer will insert or update a value in the database at a user-configurable pace. The readers will read the record till they read the last written data element, each reader reads in a period defined by the user. The different readers are scheduled uniformly within the reading period.

All this data is logged and gives possibilities to analyse when a record is visible for different users, compared to the time when the queries were started or ended.

Before the test, there are 30 000 records stored in the database, each tests takes 500 seconds and results start to be gathered after 30 seconds. In the tests is chosen to write every 0.5s and read with 5 readers every 10ms in MongoDB. In HBase there are 10 readers who read every 30ms.

@@

YCSB was deployed on a single instance and used to calibrate the load on both systems. An individual record has 10 fields which each field a size of 100 bytes. The workload existed out of 20\% inserts and updates, 40\% selects and 20\% scans of an uniform spread between 1 and 100. The order of the reading and updating of the records is defined by a Zipfian\footnote{Some record are popular, others rare to be used.} distribution. The basic load for HBase is an average of 600 queries/second spread over 50 threads, for MongoDB there are 15 threads with a total of 200 queries/second.

@@

In case of a stop of a HBase server in this configuration, the queries will halt till the lease for the region has been expired, this can take between 0 seconds and 180, depending on the moment of the action. In case of a graceful stop of the service, the service will release the lease and a faster handover is possible. During the tests it happened that with a hard stop of the service, no queries can be executed on the database connection till the server is brought back online. In this scenario it was needed to manually dis- en reconnect to the database. A reason why this happens is still unknown.

For MongoDB, there is no difference between the graceful or hard stop of an instance; in case it was a secondary, no influences on the latency will be seen. In case a primary was stopped, the data will be temporarily unavailable for a few seconds. It seems that in a hard stop, there is still a messaging of the shut down to the other nodes. In case of a network interruption, the results are varied. It goes from no influence, a temporary unavailability for a few seconds, or no completed queries the whole time the server is down. If all connections are reinitialised for the last case, new queries can executed. If this is not done the table seems empty.

@@

\subsection{Consistency}
Based on consistency, HBase and MongoDB have a different approach. In MongoDB it is possible to configure the read and write queries, in HBase there is the choice to use cache at client side. In the tests the cache is disabled but the different options of MongoDB are tested.

HBase: If a record is read while a write query is inserting or updating the value, the read query will wait till the completion of the write query and return immediately the correct value. If a read query is send too soon, it will return the old data. Each query that is finished after the completion of a write query, will return the new data. A graphical representation of this can be seen in figure \ref{fig:consistentie-hbase-voorbeeld}. The stop time of the reader follows the one of the writer, together with extra information, it can be concluded that this is coming from the blocking till the write query has finished.

@@

MongoDB: A read query can return the old or new data before the write query has been completed. Despite of the use of a readers/writer lock mechanism, it seems that after releasing the lock, the write query still has to execute extra steps. When reading from a secondary or the nearest server, there will be monotone read consistency, except if the driver of MongoDB selects another server between the reads, as there is no guarantee that the other servers already have the value. The process of finding the nearest server happens periodically in the background. \\
From the test results, the writer configuration has no influence on consistency window for the different readers if the start moments of the writers and the successful read queries are compared. The difference is the guarantee a user has when a write query has finished.

An overview of the consistency windows is shown in figure \ref{fig:consistentie-hbase-correct} and \ref{fig:consistentie-mongodb-correct} for respectively HBase and MongoDB.

\begin{figure}[!htf]

\begin{tikzpicture}

	\begin{axis}[
		xlabel={Start of the read query (ms)},
		ylabel={Percentage new value},
		domain = 0:1,
		grid = major,
		height=7cm,
		width=.45\textwidth
	]

	\addplot[color=black,mark=x]
	        plot coordinates {
	       		(0, 2.6)
				(3, 68)
				(6,90)
				(9,93)
				(12,94)
				(15,96)
				(18,96)
				(21,96)
				(24,97)
				(27,97)
	        };

	\end{axis}
\end{tikzpicture}
\caption{Consistency: Percentage of the queries started at the given time moment that will read the newest inserted data. The average latency in a random read transaction is around 6ms.}
\label{fig:consistentie-hbase-correct}
\end{figure}

\begin{figure}[!htf]

\begin{tikzpicture}

	\begin{axis}[
		xlabel={Start of the read query (ms)},
		ylabel={Percentage new value},
		domain = 0:1,
		grid = major,
		height=7cm,
		width=.45\textwidth,
		legend pos= south east,
		cycle list name=black white
	]

		\addplot
		        plot coordinates {
		       		(0, 73)
					(2, 98)
					(4,99)
					(6,99)
					(8,99)
		        };
		\addlegendentry{Primary}

		\addplot
	      plot coordinates {
		       		(0, 77)
					(2, 99)
					(4,99)
					(6,99)
					(8,100)
	      };
		\addlegendentry{Primary-preferred}

	\addplot
	        plot coordinates {
	       		(0, 26)
				(2, 77)
				(4,91)
				(6,91)
				(8,92)
	        };
	\addlegendentry{Nearest}

	\addplot
   plot coordinates {
	       		(0, 0)
				(2, 61)
				(4,82)
				(6,85)
				(8,89)
   };
	\addlegendentry{Secondary}
	\end{axis}
\end{tikzpicture}
\caption{Consistency: Percentage of the queries started at 0, 2, 4 and 8ms  that will read the newest inserted data with the each row for MongoDB. The average latency on a read transaction is for all read queries around 1ms. }
\label{fig:consistentie-mongodb-correct}
\end{figure}

\section{Future work}\label{sec:futurework}
The test executed on the different systems are a start to have a quantified approach on the consistency and availability of the different systems. Besides testing more systems, deeper testing can be done and extensions can be added to the benchmarking tool.

@@

Secondly, the third element of CAP can be included: Partition tolerance. Right now the connection nodes are chosen when setting up the test, an extension could be to test all connections and see what happens in the case of a partition tolerance. How is the availability of each node? How long is a primary in MongoDB still accepting queries after it has been cut off from the other instances?

@@

At last, the test parameters could be adapted, what happens with a different network infrastructure when the network distance is not equal any more, what happens with a higher and lower load? It could be possible to come up with a mathematical formula which could predict the average consistency window for example.

\section{Related work}\label{sec:related work}
The is only a limited amount of research executed towards the guaranties a database system provides towards consistency and availability. In recent paper (February 2014), Golab et al. states that there is only a limited amount of research done towards eventual consistency \cite{golab2014eventually}. They present the current research and describe a missing category of testing methods towards consistency. This category is called the passive analysis, this analysis focusses on how a database user observers the consistency guaranties. This is different from the active analysis where they mostly focus on the time it takes to replicated to all nodes.  With the testing method described in this paper, both are covered: as well the delays before the data is present everywhere but also on the acting of the specific systems, in example the behaviour of HBase.

Another extension of YCSB called YCSB++\cite{patil2011ycsb++}, provides more logging information on all systems in the first place, but they also test the consistency of HBase in regards of the client caches. These caches are by default enabled on the client side, but the submission of the client cache towards the system does not only depend on the time. Also the amount of traffic generated at the client influences this speed. In the study they compare different cache sizes and this shows already a difference in time. In case data needs to be strict consistent, it is better to disable caching at this moment, this way the behaviour is more predictable and in general also faster available for other clients.

@@

In the availability of the systems, HBase is in the standard configuration longer unavailable but with a change in the session time out, they can be made more the same. In some occasions both systems have a strange behaviour of not returning any results, in the case of HBase this is for a hard stop, in the case of MongoDB it is for a network partition. This doesn't happen in the other situations and in case of reconnecting to the database, the problems are solved. 
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix
